Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-10-01 paper/tidy-missing-data-paper.Rmd
fixing a couple more table captions

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
  \usepackage{float}
  \usepackage{booktabs}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
# bibliography: tidy-missing-data-paper.bib
bibliography: "tidy-missing-data-paper-no-jss.bib"
cls: jss.cls
---

@@

airquality %>%
  group_by(Month) %>%
  miss_var_summary() %>%
  ungroup() %>%
  slice(1:10) %>%
  knitr::kable(caption = "\\texttt{miss\\char`_var\\char`_summary} combined with \\texttt{group\\char`_by} provides a grouped summary of the missingness in each variable, for each Month of the airquality dataset. Only the first 10 rows are shown. There are more ozone missings in June than May.",
               digits = 1)
               # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

knitr::kable(list(t1, t2),
             caption = "Tabulating missingness for variables (left) and cases (right) to understand missingness patterns. 14 variables have 0 - 3 missings, 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missings. About 30\\% of cases have no missings, 45\\% of cases have 1 - 6 missings, and about 23\\% of cases have 8 or more missings. There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.",
             digits = 1)
             # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

The *nabular* data structures discussed in the paper are simple by design, to promote flexibility. They could be used to create different visualizations than were shown in the paper. The analyst can use the data structures to decide on appropriate visualization for their problem. The data structures could also be used to support interactive graphics, in the manner of MANET and ggobi. Linking the plots (via linked brushing) could facilitate exploration of missingness, and could be implemented with \pkg{plotly} [@plotly] for added interactivity. Animating between different sets of imputed values could also be possible with packages like \pkg{gganimate} [@gganimate].

>>>>

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
  \usepackage{float}
  \usepackage{booktabs}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
bibliography: "tidy-missing-data-paper.bib"
cls: jss.cls
---

@@

airquality %>%
  group_by(Month) %>%
  miss_var_summary() %>%
  ungroup() %>%
  slice(1:10) %>%
  knitr::kable(caption = "Output of \\code{airquality \\%>\\% group\\char`_by(Month) \\%>\\% miss\\char`_var\\char`_summary()} provides a grouped summary of the missingness in each variable, for each month of the airquality dataset. Only the first 10 rows are shown. There are more ozone missings in June than May.",
               digits = 1)
               # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

knitr::kable(list(t1, t2),
             caption = "Summary tables to help understand missingness patters. Output of \\code{miss\\_var\\_table(housing)} (left), tabulating missingness for variables, and output of \\code{miss\\_case\\_table(housing)}. There are 14 variables with 0 - 3 missings, 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missings. About 30\\% of cases have no missings, 45\\% of cases have 1 - 6 missings, and about 23\\% of cases have 8 or more missings. There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.",
             digits = 1)
             # booktabs = TRUE)
  # kable_styling(latex_options = c("hold_position"))

@@

The *nabular* data structures discussed in the paper are simple by design, to promote flexibility. They could be used to create different visualizations than were shown in the paper. The analyst can use the data structures to decide on appropriate visualization for their problem. The data structures could also be used to support interactive graphics, in the manner of \proglang{MANET} and \proglang{ggobi}. Linking the plots (via linked brushing) could facilitate exploration of missingness, and could be implemented with \pkg{plotly} [@plotly] for added interactivity. Animating between different sets of imputed values could also be possible with packages like \pkg{gganimate} [@gganimate].
